Analyzing the Effect of Surface in Tennis Grand Slams

### 2018-09-14 contest/surface-type/paper-tennis.Rmd
spell check lol

---
title: "Analyzing the Effect of Surface in Tennis Grand Slams"
author: "Kayla Frisoli, Shannon Gallagher, and Amanda Luby"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{hyperref}
abstract: "Tennis grand slams consist of the Australian Open, French Open, Wimbledon, and US Open, which are played on hard (Plexicushion), clay, grass, and hard (DecoTurf) courts, respectively.  The surface type may substantially impact ball speed, height, and spin as well as player speed and agility.  It is also believed that play style and practice habits may contribute to different results across surface types.  For example, Rafael Nadal is thought to be the best clay court player of all time whereas Roger Federer is particularly known for dominance at Wimbledon.  On the women's side, Serena Williams once struggled on clay courts but has seemingly transformed her style to perform better on clay courts, but has perhaps suffered on grass as a consequence.  In this analysis, we examine the result of the top 100 players in grand slams from 2013-2017 across the four different surfaces.  We create a hierarchical model with fixed and random effects to predict the number of points won in a match.  We take into consideration player-specific effects, nationality (which is thought to have an effect on play style), sex, ranking, ELO, and game statistics.   We assess the fit of our model using standard statistical techniques (e.g. MSE, AIC, BIC, residual diagnostics) in addition to `common knowledge' factors (for instance, Rafael Nadal should be indicated as a superior clay court player by the model).  We compare the results of top 100 players across grand slams to examine the effect of court surface. We also provide an in-depth analysis of Nadal, Federer, and S. Williams."
bibliography: master.bib
---

@@

We first examine the distribution of points earned per match. We find that the distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams. As expected, there are more points earned in the WTA than the ATP due to the differing numbers of games played. Also unsurprisingly, the winners of the match tended to earn more points than the losers.

@@

```{r, fig.height=4, fig.width=12}
france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))

@@

@@

p3 <- ggplot(aus_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
  geom_line() + 
  # geom_point() +
  labs(title = "Australian players",
       x = "", y = "Win proportion") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme

grid.arrange(p1,p2, p3, nrow=1, widths = c(.29, .29, .42),
             top = textGrob("Home court advantage debunked: win proportion by country",
                            gp=gpar(fontsize=20, fontfamily="serif")),
             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size as the tournament goes on",
                           # x = 0,
                           # y = 0.5,
                           just = "left"))

@@

@@

p1 <- ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
  geom_line() + 
  labs(title = "Win proportion among Spaniards",
       # caption = "Note: we only look at the first 4 rounds due to decreasing sample size",
       x = "", y = "Win proportion") + 
  scale_color_manual(guide=FALSE, values=tournament_colors) + my_theme  
  # geom_point() + 
  # labs(title = "Proportion of wins among Spaniards") +
  # geom_errorbar(aes(ymin = win_prop - win_prop_se,
  #                   ymax = win_prop + win_prop_se),
  #               width=.2,
  #                position=position_dodge(0.05))

@@

grid.arrange(p1,p2, nrow=1, widths = c(.44, .56),
             top = textGrob("Spaniards perform better on Clay: debunked",
                            gp=gpar(fontsize=20)),
             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
                           # x = 0,
                           # y = 0.5,
                           just = "left",
                           gp=gpar(fontfamily="serif")))

>>>>

---
title: "Analyzing the Effect of Surface in Tennis Grand Slams"
author: "###########"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{hyperref}
abstract: "Tennis grand slams consist of the Australian Open, French Open, Wimbledon, and US Open, which are played on hard (Plexicushion), clay, grass, and hard (DecoTurf) courts, respectively.  The surface type may substantially impact ball speed, height, and spin as well as player speed and agility.  It is also believed that play style and practice habits may contribute to different results across surface types.  For example, Rafael Nadal is thought to be the best clay court player of all time whereas Roger Federer is particularly known for dominance at Wimbledon.  On the women's side, Serena Williams once struggled on clay courts but has seemingly transformed her style to perform better on clay courts, but has perhaps suffered on grass as a consequence.  In this analysis, we examine the result of the top 100 players in grand slams from 2013-2017 across the four different surfaces.  We create a hierarchical model with fixed and random effects to predict the number of points won in a match.  We take into consideration player-specific effects, nationality (which is thought to have an effect on play style), sex, ranking, ELO, and game statistics.   We assess the fit of our model using standard statistical techniques (e.g. MSE, AIC, BIC, residual diagnostics) in addition to `common knowledge' factors (for instance, Rafael Nadal should be indicated as a superior clay court player by the model).  We compare the results of top 100 players across grand slams to examine the effect of court surface. We also provide an in-depth analysis of Nadal, Federer, and S. Williams."
bibliography: master.bib
---

@@

We first examine the distribution of points earned per match to assess normality. The distribution of points earned is approximately normal. This distribution is similar across tournament, with Wimbledon differing slightly from the other grand slams. As expected, there are more points earned in the WTA than the ATP due to the differing numbers of games played. Also unsurprisingly, the winners of the match tended to earn more points than the losers.

@@

```{r, fig.height=4, fig.width=14, fig.align='center'}
france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))

@@

gbr_wins <- filter(gs_players, ioc == "GBR", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))

@@

p3 <- ggplot(aus_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
  geom_line() + 
  # geom_point() +
  labs(title = "Australian players",
       x = "", y = "Win proportion") + 
  scale_color_manual(guide=FALSE, values=tournament_colors) + my_theme

p4 <- ggplot(gbr_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
  geom_line() + 
  # geom_point() +
  labs(title = "British players",
       x = "", y = "Win proportion") + 
  scale_color_manual(" ", values=tournament_colors) + my_theme

grid.arrange(p3, p1, p2,  p4, nrow=1, widths = c(.22, .22, .22, .34),
             top = textGrob("Home court advantage: debunked",
                            gp=gpar(fontsize=20, fontfamily="serif")),
             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
                           # x = 0,
                           # y = 0.5,
                           just = "left",
                           gp=gpar(fontfamily="serif")))

@@

```{r, fig.height=4, fig.width=12, fig.align='center'}

@@

p1 <- ggplot(spain_wins, aes(round, win_prop, color=tournament, group=tournament)) + 
  geom_line() + 
  labs(title = "Win proportion among Spaniards",
       # caption = "Note: we only look at the first 4 rounds due to decreasing sample size",
       x = "", y = "Win proportion") + 
  scale_color_manual(guide=FALSE, values=tournament_colors) + my_theme

@@

grid.arrange(p1, p2, nrow=1, widths = c(.44, .56),
             top = textGrob("Spaniards perform better on Clay: debunked",
                            gp=gpar(fontsize=20, fontfamily="serif")),
             bottom = textGrob("Note: we only look at the first 4 rounds due to decreasing sample size",
                           # x = 0,
                           # y = 0.5,
                           just = "left",
                           gp=gpar(fontfamily="serif")))
